---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data
Before loading the file, the provided zip-archive is unzipped and the containing file is stored in the same directory. Then, the file CSV file is loaded.

```{r}
data <- read.csv("activity.csv", header=T)
data <- cbind(data, date_correct = as.Date(data$date, format="%Y-%m-%d"))
data <- subset(data, select=c(steps,date_correct,interval))
colnames(data) <- c("steps","date","interval")
```

The date is not recognized as date. Therefore, we convert it to the type date using the mentioned functions.


## What is mean total number of steps taken per day?

We want to calculate the total number of steps per day. The data should be displayed as a histogram. We will use the library ggplot2 to create the relevant plot.

```{r}
library(ggplot2)

totals <- aggregate (x = data$steps, by=list(data$date), FUN = sum)

ggplot(totals, aes(x=x)) + geom_histogram(binwidth=1000)
```

In addition, we want to show the mean and the median per day.
```{r}
means <- aggregate (x = data$steps, by=list(data$date), FUN = mean)
medians <- aggregate (x = data$steps, by=list(data$date), FUN = median)

means
medians
```



## What is the average daily activity pattern?

To obtain the average daily activity pattern, we first calculate the menas by interval. As there are a lot of NA values, we remove them in order to get a valid result. Then, we plot the data.
```{r}
means_plot <- aggregate(x = data$steps, by=list(data$interval), FUN = mean, na.rm=TRUE, na.action=na.omit)
p <- ggplot(means_plot, aes(x=Group.1, y = x))+geom_line() + xlab("")
p
```
Next, we determine the interval with the maximum average number of steps.
```{r}
max_steps = means_plot[means_plot$x == max(means_plot$x),]
colnames(max_steps) = c("Interval", "Mean no. of steps")
max_steps
```
The relevant interval is `r max_steps$Interval`.

## Imputing missing values

First, we get the number of missing values.
```{r}
sum(is.na(data$steps))
```

To make up for the missing values, we pursue the strategy of imputing the mean number of steps per interval. The new data set is denoted as `data_new`.
```{r}
require(Hmisc)
require(plyr)
data_new <- ddply(data, "interval", mutate, imputed.value=impute(steps, mean))
data_new$steps <- floor(data_new$imputed.value)
data_new <- subset(data_new, select=c(steps,date,interval))
```

Next, we will plot the new data.

```{r}

totals_new <- aggregate (x = data$steps, by=list(data$date), FUN = sum)

ggplot(totals_new, aes(x=x)) + geom_histogram(binwidth=1000)
```

Lastly, we will calculate the mean and the median.

```{r}
means <- aggregate (x = data_new$steps, by=list(data_new$date), FUN = mean)
medians <- aggregate (x = data_new$steps, by=list(data$date), FUN = median)

means
medians
```

As we are using the means by group, the means do not change. However, when considering different ways of aggregation, this may change.

## Are there differences in activity patterns between weekdays and weekends?

First, we determine the day of the week of the given date. As this R installation is in German, I use the German words for Saturday and Sunday respectively. Then we plot the graph according to the instructions.
```{r}
data_new$weekdays <- weekdays(data_new$date, abbreviate=FALSE)
data_new$weekend_indicator <- ifelse(data_new$weekdays == "Samstag" | data_new$weekdays == "Sonntag", "weekend", "weekday")
data_new <- subset(data_new, select=c(steps,date,interval,weekend_indicator))

means_plot <- aggregate(x = data_new$steps, by=list(data_new$interval, data_new$weekend_indicator), FUN = mean, na.rm=TRUE, na.action=na.omit)
ggplot(means_plot, aes(x=Group.1, y = x))+geom_line() + xlab("")+facet_wrap(~Group.2, nrow=2)
```

There seems to be a difference between weekdays and weekend. For example, weekday activity starts earlier but also ends ealier, while weekend activity keeps on going for later intervals.
